\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Model

### 2020-05-06 Tex/appendix.tex
working to trim

>>>>

\section{c-ERGM Estimation}

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the c-ERGM.

The common alternative relies on Monte Carlo methods to approximate the normalizing constant by simulating a large set of networks \citep{hunter2006inference,hummel2012improving}. The resulting estimator, the Monte Carlo MLE (MCMLE), is approximately consistent, meaning that it converges to the MLE as the sample size, i.e., the number of simulated networks, increases. However, one drawback is that with the number of nodes in the Supreme Court citation network being in the order of $10,000$, obtaining the MCMLE is computationally expensive \citep{schmid2017exponential} and the success of the algorithm heavily relies on the starting parameter vector $\theta_0$, which is ideally chosen in the proximity of the unknown MLE \citep{hummel2012improving}. The prevailing choice for $\theta_0$ is the maximum pseudo-likelihood estimation (MPLE) \citep{strauss1990pseudolikelihood}, a fast estimation method that is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form. The MPLE is simple to obtain, but does not guarantee a starting value close to the MLE \citep{SchmidHunter2020}.

We estimate the MCMLE for every term of the c-ERGM between 1950 and 2015. While the network for the 1950 term consists of $1962$ nodes, the size of the network monotonically increases to $10,020$ nodes for the 2015 term, making estimation more challenging as the network grows. The growth in the network also serves as our reason for estimating a separate model for each term. First, the network is large, with each term introducing the potential for hundreds-of-thousands of new citations. This means we have the statistical power necessary to estimate a separate set of parameter values for each term. Second, based on the findings of \citet{shalizi2013consistency}, we know that the set of parameter values that best fits a network should change as more nodes are added to the network. This means it would be inappropriate to assume a constant set of parameter values as more cases are added each term.

The MCMLE of networks up until the 90s was obtainable in a reasonable time frame starting at the MPLE and sampling $10,000$ networks to approximate the normalizing constant. However, the estimation of most networks in the 90s with the MPLE as starting values was not feasible in an reasonable time frame anymore. Instead, we improved the choice of starting value $\theta_0$ by fixing it at the MCMLE of the previous term $t-1$ and successfully obtained the MCMLE of the network at term $t$. But even this approach started to fail for networks around the turn of the millennium. Neither the MPLE nor the MCMLE of previous terms as starting values led to successful estimation, and neither did the Stepping algorithm \citep{hummel2012improving}. 
The MCMLE for these large citation networks was obtained by setting the starting value according to an approach introduced by \citet{SchmidHunter2020}. This method is based on the fact that the MLE of exponential family distributions is solely a function of the vector of sufficient statistics $h(C_t, C_{<t})$, meaning that the MLE of two networks $A$ and $B$ is equal if $h(A)=h(B)$. However, the MPLE of networks with the same sufficient statistics is not necessarily the same. Instead of starting the MCMLE algorithm at the network's MPLE, \citet{SchmidHunter2020} propose searching for a network $C_t^*$ using simulated annealing \citep{Kirkpatrick83}, with $h(C_t, C_{<t}) = h(C_t^*, C_{<t})$ and a weak dependence structure among unfixed ties. The MPLE of such a network is similar to the MLE, making it an effective starting value for the MCMLE algorithm.

### 2020-05-06 Tex/paper.tex
working to trim

@@

\usepackage{geometry}
\usepackage{float}

@@

@@

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 1cm 0cm 2cm,clip=true ]{images/citations_recip.pdf}
\caption{Illustration of a reciprocal tie between two US Supreme Court opinions.}
\label{fig:reciprocity}
\end{figure}

@@

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 3cm 6cm 8cm,clip=true ]{images/citations_pop.pdf}
\caption{Illustration of ties sent to a landmark Supreme Court opinion via citations.}
\label{fig:popularity}
\end{figure}

@@

\subsection{c-ERGM definition}

@@

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 0cm 0cm 0cm,clip=true  ]{images/daggish.pdf}
\caption{Illustration of temporal structure of the Supreme Court Citation Network. $C_{\leq t}$ is the entire set of citations (and non-citations) on which citations and non-citations at time $t$ (i.e., $C_t$) depend. $C_t$ are conditioned on the citations and non-citations established before time $t$ (i.e., $C_{< t}$). The shaded small squares are hypothetical observed citations, and the white small squares are citations that could have been observed but were not. The regions of the citation matrix that are represented by large white rectangles are citations that could not have been observed since the citing case would have been decided in a term that preceded the term of the cited case. The citing case ID is given in the row and the prospective cited case is given in the column. }
\label{fig:ctergm}
\end{figure}

@@

Though the likelihood function of the c-ERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increases by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$). The logit form conditional probability is well known for the ERGM family \citep{goodreau2009birds}.

We should note that there is nothing about the c-ERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, the term for that dynamic would indicate that there is an increase in the log odds that case $i$ cites case $j$ with each other case that has cited case $j$.  That finding could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case---as in the Baker v. Carr passage cited above. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-ERGM can be specified to account for the dependence in the data.

\subsection{c-ERGM Estimation}

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the c-ERGM.

The common alternative relies on Monte Carlo methods to approximate the normalizing constant by simulating a large set of networks \citep{hunter2006inference,hummel2012improving}. The resulting estimator, the Monte Carlo MLE (MCMLE), is approximately consistent, meaning that it converges to the MLE as the sample size, i.e., the number of simulated networks, increases. However, one drawback is that with the number of nodes in the Supreme Court citation network being in the order of $10,000$, obtaining the MCMLE is computationally expensive \citep{schmid2017exponential} and the success of the algorithm heavily relies on the starting parameter vector $\theta_0$, which is ideally chosen in the proximity of the unknown MLE \citep{hummel2012improving}. The prevailing choice for $\theta_0$ is the maximum pseudo-likelihood estimation (MPLE) \citep{strauss1990pseudolikelihood}, a fast estimation method that is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form. The MPLE is simple to obtain, but does not guarantee a starting value close to the MLE \citep{SchmidHunter2020}.

We estimate the MCMLE for every term of the c-ERGM between 1950 and 2015. While the network for the 1950 term consists of $1962$ nodes, the size of the network monotonically increases to $10,020$ nodes for the 2015 term, making estimation more challenging as the network grows. The growth in the network also serves as our reason for estimating a separate model for each term. First, the network is large, with each term introducing the potential for hundreds-of-thousands of new citations. This means we have the statistical power necessary to estimate a separate set of parameter values for each term. Second, based on the findings of \citet{shalizi2013consistency}, we know that the set of parameter values that best fits a network should change as more nodes are added to the network. This means it would be inappropriate to assume a constant set of parameter values as more cases are added each term.

The MCMLE of networks up until the 90s was obtainable in a reasonable time frame starting at the MPLE and sampling $10,000$ networks to approximate the normalizing constant. However, the estimation of most networks in the 90s with the MPLE as starting values was not feasible in an reasonable time frame anymore. Instead, we improved the choice of starting value $\theta_0$ by fixing it at the MCMLE of the previous term $t-1$ and successfully obtained the MCMLE of the network at term $t$. But even this approach started to fail for networks around the turn of the millennium. Neither the MPLE nor the MCMLE of previous terms as starting values led to successful estimation, and neither did the Stepping algorithm \citep{hummel2012improving}. 
The MCMLE for these large citation networks was obtained by setting the starting value according to an approach introduced by \citet{SchmidHunter2020}. This method is based on the fact that the MLE of exponential family distributions is solely a function of the vector of sufficient statistics $h(C_t, C_{<t})$, meaning that the MLE of two networks $A$ and $B$ is equal if $h(A)=h(B)$. However, the MPLE of networks with the same sufficient statistics is not necessarily the same. Instead of starting the MCMLE algorithm at the network's MPLE, \citet{SchmidHunter2020} propose searching for a network $C_t^*$ using simulated annealing \citep{Kirkpatrick83}, with $h(C_t, C_{<t}) = h(C_t^*, C_{<t})$ and a weak dependence structure among unfixed ties. The MPLE of such a network is similar to the MLE, making it an effective starting value for the MCMLE algorithm.

@@

\begin{figure}[htp]
\includegraphics[scale=0.5]{degree_distribution}
\caption{The in- and outdegree distribution of the Supreme Court Citation Network from 1937 - 2001. There are cases with an indegree >50, but they are not captured in this figure.}
 \label{degree_dist}
\vspace{-.25cm}
\end{figure}

\begin{figure}[htp]
\includegraphics[scale=0.35,clip=true,trim=.5cm 0cm 0cm 2cm]{images/citations1}
\caption{ Supreme Court Citation Network, 1937-2001. Nodes are Supreme Court cases, with size based on incoming citations. Salient cases ({\em Oxford Guide}) are in yellow. Each network layer contains cases decided under a different chief justice. Because our data is temporally bound, cases for the Hughes and Rehnquist courts are not complete. }
 \label{fig:networkviz}
\vspace{-.25cm}
\end{figure}

@@

  \begin{figure}[t]
\includegraphics[width=15cm]{SCC_AIC_BIC}
\caption{AIC and BIC for the full and the independent model for the time frame 1950-2015 }
 \label{AIC_BIC}
\vspace{-.25cm}
\end{figure}

@@

We recognize two important limitations of the current analysis, which should be addressed in future iterations of this work or in future studies. First, our theoretical claims and analyses treat all citations---positive and negative---as equivalent. Both data and methodological limitations will need to be overcome to consider these two citation types. Second, there is nothing in the c-ERGM that governs the generation of new cases (i.e., cases are assumed to arise exogenously, unrelated to the history of the citation network). As new cases are often taken up by the Supreme Court in order to test, clarify, and/or challenge existing precedents, it is likely inappropriate to assume that new cases arise exogenous to the existing citation network. Overcoming this limitation would require methodological innovation in the c-ERGM as well as theoretical development regarding exactly how citation structure might predict case emergence.

\newpage

>>>>

\usepackage[moderate]{savetrees}

@@

 \usepackage{float}

@@

\usepackage{geometry}

@@

\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 1cm 0cm 2cm,clip=true ]{images/citations_recip.pdf}
\caption{Illustration of a reciprocal tie between two US Supreme Court opinions.}
\label{fig:reciprocity}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 3cm 6cm 8cm,clip=true ]{images/citations_pop.pdf}
\caption{Illustration of ties sent to a landmark Supreme Court opinion via citations.}
\label{fig:popularity}
\end{figure}

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 0cm 0cm 0cm,clip=true  ]{images/daggish.pdf}
\caption{Illustration of temporal structure of the Supreme Court Citation Network. $C_{\leq t}$ is the entire set of citations (and non-citations) on which citations and non-citations at time $t$ (i.e., $C_t$) depend. $C_t$ are conditioned on the citations and non-citations established before time $t$ (i.e., $C_{< t}$). The shaded small squares are hypothetical observed citations, and the white small squares are citations that could have been observed but were not. The regions of the citation matrix that are represented by large white rectangles are citations that could not have been observed since the citing case would have been decided in a term that preceded the term of the cited case. The citing case ID is given in the row and the prospective cited case is given in the column. }
\label{fig:ctergm}
\end{figure}

@@

Though the likelihood function of the c-ERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increases by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$). The logit form conditional probability is well known for the ERGM family \citep{goodreau2009birds}. Estimation of the c-ERGM parameters, which is done with a computationally-intensive, simulation-based approach, is presented in detail in the appendix.

%We should note that there is nothing about the c-ERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, the term for that dynamic would indicate that there is an increase in the log odds that case $i$ cites case $j$ with each other case that has cited case $j$.  That finding could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case---as in the Baker v. Carr passage cited above. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-ERGM can be specified to account for the dependence in the data. Estimation of the c-ERGM parameters, which is done with a computationally-intensive, simulation-based approach, is presented in detail in the appendix.

@@

\begin{figure}[H]
\includegraphics[scale=0.5]{degree_distribution}
\caption{The in- and outdegree distribution of the Supreme Court Citation Network from 1937 - 2001. There are cases with an indegree  $>$50, but they are not captured in this figure.}
 \label{degree_dist}
\vspace{-.25cm}
\end{figure}

\begin{figure}[H]
\includegraphics[scale=0.35,clip=true,trim=.5cm 0cm 0cm 2cm]{images/citations1}
\caption{ Supreme Court Citation Network, 1937-2001. Nodes are Supreme Court cases, with size based on incoming citations. Salient cases ({\em Oxford Guide}) are in yellow. Each network layer contains cases decided under a different chief justice. Because our data is temporally bound, cases for the Hughes and Rehnquist courts are not complete. }
 \label{fig:networkviz}
\vspace{-.25cm}
\end{figure}

@@

  \begin{figure}[H]
\includegraphics[width=15cm]{SCC_AIC_BIC}
\caption{AIC and BIC for the full and the independent model for the time frame 1950-2015 }
 \label{AIC_BIC}
\vspace{-.25cm}
\end{figure}

@@

%We recognize two important limitations of the current analysis, which should be addressed in future iterations of this work or in future studies. First, our theoretical claims and analyses treat all citations---positive and negative---as equivalent. Both data and methodological limitations will need to be overcome to consider these two citation types. Second, there is nothing in the c-ERGM that governs the generation of new cases (i.e., cases are assumed to arise exogenously, unrelated to the history of the citation network). As new cases are often taken up by the Supreme Court in order to test, clarify, and/or challenge existing precedents, it is likely inappropriate to assume that new cases arise exogenous to the existing citation network. Overcoming this limitation would require methodological innovation in the c-ERGM as well as theoretical development regarding exactly how citation structure might predict case emergence.
